---
title: "Introduction to ggInterval"
subtitle: "ggInterval: an R package for visualizing interval-valued data using ggplot2" 
author: "Bo-Syue Jiang and Han-Ming Wu*"
date: "June 30, 2025"
output: 
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
    number_sections: true  
vignette: >
  %\VignetteIndexEntry{Introduction to dataSDA}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(out.width = "100%")
knitr::opts_chunk$set(fig.align = 'center')
library(knitr)
library(ggInterval)
library(RSDA)
```

$$\\[0.1in]$$ Exploratory data analysis (EDA) remains indispensable for revealing a dataset's salient features prior to formal modeling. Classic EDA relies on graphical summaries such as boxplots, histograms, and scatterplots, supplemented by dimension-reduction tools and interactive displays. Yet modern data are increasingly high-volume and structurally rich: single observations may be recorded not as scalars but as intervals, histograms, or full empirical distributions, collectively known as symbolic data. Conventional graphics cannot natively accommodate these objects, creating a gap between contemporary data structures and available visualization tools. To bridge this gap, we introduce <b>ggInterval</b> (formerly <b>ggESDA</b>*), an <b>R</b> package for visualizing interval-valued symbolic data. The package delivers seamless integration with the <b>ggplot2</b> workflow. We detail the package design, explain its extensibility, and illustrate its practical value through two real-world symbolic datasets. The results demonstrate that <b>ggInterval</b> enables analysts to uncover structure and anomalies in interval-valued data that would remain hidden with traditional scalar-oriented graphics, thereby enhancing the EDA toolkit for modern applied statistics and data science. The <b>ggInterval</b> package is currently available on the Comprehensive R Archive Network (CRAN) at <https://CRAN.R-project.org/package=ggInterval> and <https://hmwu.idv.tw/ggInterval/>.

* **Current version** (build on June 30, 2025): [ggInterval_0.2.2.zip](https://hmwu.idv.tw/ggInterval/ggInterval_0.2.2.zip)

* **Citation**: Jiang, B.S. and Wu, H.M.\* (2025), ggInterval: an R package for visualizing interval-valued data using ggplot2. R package version 0.2.2, https://CRAN.R-project.org/package=ggInterval. 

* **Vignettes**: [ggInterval_Jiang&Wu.pdf](https://hmwu.idv.tw/ggInterval/ggInterval_Jiang&Wu.pdf)

<font size = "0.5"><em> * The package is renamed to better reflect the focus of our work on interval-valued data visualization. </em></font>


$$\\[0.1in]$$ 
 

