---
title: "Introduction to ggInterval"
subtitle: "A source-portable companion for interval-valued visualization"
author: "Bo-Syue Jiang and Han-Ming Wu"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to ggInterval}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)

library(ggInterval)
library(ggplot2)
```

# Overview

`ggInterval` extends `ggplot2` for interval-valued data. The package supports
visualizations such as scatter plots, histograms, index plots, radar plots,
line plots, PCA displays, and correlation heatmaps, while keeping the
layered grammar familiar to `ggplot2` users.

This vignette focuses on source-portable examples that can be rebuilt directly
from the package. It uses built-in data, the bundled `data.csv` file for the
aggregation example, and optional sections for packages such as `HistDAWass`
and `MAINT.Data`.

# Built-in Data

```{r built_in_data}
data(facedata)
data(Environment)

facedata
Environment

Subjects <- substr(rownames(facedata), 1, 3)
Subjects
```

# Basic Usage

```{r ggplot2_workflow}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, col = Species)) +
  geom_point() +
  scale_color_manual(values = c("red", "blue", "goldenrod"))

ggInterval_scatterplot(
  facedata,
  aes(x = AD, y = BC, fill = Subjects),
  col = "black"
) +
  scale_fill_brewer(palette = "Set1") +
  labs(fill = "Subjects")
```

# Aggregating Classical Data into Interval-Valued Data

```{r aggregation_examples}
breastData <- read.csv("data.csv")
breastData$X <- NULL
breastData <- dplyr::select(breastData, -id)

breastData_kmeans <- classic2sym(breastData, groupby = "kmeans", k = 5)
head(breastData_kmeans$intervalData[, 1:4])

breastData_diag <- classic2sym(breastData, groupby = "diagnosis")
head(breastData_diag$intervalData[, 1:4])
```

# Descriptive Statistics

```{r descriptive_statistics}
mean(facedata)
sd(facedata)

cov(facedata$AD, facedata$BC, method = "BD")
cor(facedata$AD, facedata$BC, method = "BD")

summary(facedata$AD)
summary(Environment[1:3, ])
```

# Univariate Plots

## Index Plots and Index Images

```{r index_plots}
ggInterval_indexplot(facedata, aes(x = AD))

ggInterval_indexplot(facedata, aes(x = AD, fill = Subjects)) +
  scale_fill_brewer(palette = "Set1") +
  geom_vline(xintercept = mean(facedata$AD), color = "darkgray") +
  labs(fill = "Subjects")

ggInterval_indexImage(facedata, aes(x = AD)) +
  coord_flip()

ggInterval_indexImage(facedata, aes(x = AD), full_strip = TRUE) +
  coord_flip()
```

```{r index_plot_ordering}
face.ip <- ggInterval_indexplot(facedata, aes(fill = Subjects), plotAll = TRUE) +
  scale_fill_brewer(palette = "Set1") +
  labs(x = "", y = "", fill = "Subjects")
face.ip

b <- ggInterval_indexplot(facedata, aes(fill = Subjects),
  plotAll = TRUE, row_order = "c") +
  scale_fill_brewer(palette = "Set1") +
  labs(x = "", y = "", fill = "Subjects")
b

cc <- ggInterval_indexplot(facedata, aes(fill = Subjects),
  plotAll = TRUE, row_order = "r") +
  scale_fill_brewer(palette = "Set1") +
  labs(x = "", y = "", fill = "Subjects")
cc
```

In the ordered `plotAll` displays, the outer y-axis labels are suppressed by
default. Here `row_order = "c"` reorders observations by interval centers
within each variable panel, while `row_order = "r"` reorders them by interval
ranges within each variable panel. Set `labels = TRUE` if you want the row
labels repeated inside each variable panel.

## Boxplots and Histograms

```{r boxplots_histograms}
ggInterval_boxplot(facedata, aes(x = AD), width_type = "violin-like")

ggInterval_boxplot(facedata, aes(x = AD), width_type = "quantile-depth")

ggInterval_boxplot(facedata, aes(x = AD), width_type = "side-by-side")

ggInterval_hist(facedata, aes(x = AD), method = "equal-bin", bins = 10)

ggInterval_hist(facedata, aes(x = AD), method = "unequal-bin")
```

## Line Plot

```{r lineplot_stocklike}
if (!requireNamespace("TTR", quietly = TRUE)) {
  stop("Package 'TTR' is required for this example.")
}

data("ttrc", package = "TTR")
stock.data <- subset(
  ttrc[, c("Date", "Close", "Low", "High")],
  format(Date, "%Y-%m") %in% c("1985-01", "1985-02", "1985-03")
)
stock.data$Month <- factor(
  month.abb[as.integer(format(stock.data$Date, "%m"))],
  levels = month.abb[1:3]
)
stock.data$Day <- as.integer(format(stock.data$Date, "%d"))

stock.data.LH <- classic2sym(
  stock.data,
  groupby = "customize",
  minData = stock.data$Low,
  maxData = stock.data$High
)

ggInterval_lineplot(stock.data.LH, aes(y = V1, x = Day), barWidth = 0.6) +
  geom_point(aes(y = Close), shape = 21, fill = "#D95F02",
             color = "black", size = 1.6, stroke = 0.2) +
  coord_cartesian(xlim = c(1, 31), expand = FALSE) +
  facet_wrap(~Month, ncol = 1, scales = "free_y") +
  scale_x_continuous(breaks = c(1, 8, 15, 22, 29)) +
  labs(x = "Day of month", y = "Price") +
  ggthemes::theme_economist() +
  theme(strip.text = element_text(face = "bold"))
```

## Min-Max Plot

```{r minmax_plot}
mm.plot <- function(x) {
  plot.var <<- names(facedata)[x]
  ggInterval_MMplot(facedata, aes(facedata[[plot.var]], size = 2)) +
    coord_fixed(ratio = 1) +
    theme(legend.position = "none")
}

mm.plot.list <- lapply(seq_along(facedata), mm.plot)
gridExtra::marrangeGrob(mm.plot.list, nrow = 2, ncol = 3, top = "")
```

## Center-Range Plot

```{r centerrange_plot}
ggInterval_CRplot(facedata, aes(size = 1.5), plotAll = TRUE)

facedata.scale <- scale(facedata)$intervalData
ggInterval_CRplot(facedata.scale, aes(size = 1.5), plotAll = TRUE)
```

# Bivariate Plots

## Scatter Plot with Label Controls

```{r scatterplot_labels}
ggInterval_scatterplot(
  facedata,
  aes(x = BC, y = AD, fill = Subjects),
  showLabels = TRUE,
  labelSize = 2.6,
  labelPosition = "topright",
  labelNudgeX = -0.2,
  labelNudgeY = 0.15,
  checkOverlap = FALSE,
  color = "black"
) +
  scale_fill_brewer(palette = "Set1") +
  scale_x_continuous(expand = expansion(mult = c(0.08, 0.12))) +
  scale_y_continuous(expand = expansion(mult = c(0.06, 0.10))) +
  labs(fill = "Subjects")
```

## 2D Histograms

```{r hist2d_equal}
face.2dh.equal <- ggInterval_2Dhist(
  facedata,
  aes(x = BC, y = AD, col = "white"),
  method = "equal-bin",
  xBins = 10,
  yBins = 10,
  display = "p",
  palette = "Blues",
  direction = 1,
  cell_labels = TRUE
)$plot

face.2dh.equal +
  coord_fixed(ratio = 1)
```

```{r hist2d_unequal}
face.2dh.unequal <- ggInterval_2Dhist(
  facedata,
  aes(x = BC, y = AD, col = "white"),
  method = "unequal-bin",
  display = "p",
  palette = "Blues",
  direction = 1,
  tau = 0.5
)$plot

face.2dh.unequal +
  coord_fixed(ratio = 1)
```

# Matrix Displays

## Scatter Matrix

```{r scatter_matrix}
ggInterval_scatterMatrix(facedata, aes(fill = "steelblue", alpha = 0.3))
```

## 2D Histogram Matrices

```{r hist2d_matrix_equal}
ggInterval_2DhistMatrix(
  facedata,
  aes(col = "white"),
  method = "equal-bin",
  xBins = 10,
  yBins = 10,
  display = "p",
  palette = "Blues",
  direction = 1,
  cell_labels = TRUE
)
```

```{r hist2d_matrix_unequal}
ggInterval_2DhistMatrix(
  facedata,
  aes(col = "white"),
  method = "unequal-bin",
  display = "p",
  palette = "Blues",
  direction = 1,
  cell_labels = TRUE,
  tau = 0.5
)
```

## Image Plots

```{r image_plots}
ggInterval_indexImage(
  facedata,
  plotAll = TRUE,
  full_strip = TRUE,
  column_condition = FALSE
) +
  scale_colour_distiller(palette = "Blues", direction = 1)

ggInterval_indexImage(
  facedata,
  plotAll = TRUE,
  full_strip = FALSE
) +
  scale_colour_distiller(palette = "Spectral")
```

## Radar Plots

```{r radar_plots}
library(ggthemes)

ggInterval_radarplot(
  Environment,
  plotPartial = c(4, 6),
  showLegend = FALSE,
  base_circle = FALSE,
  base_lty = 1,
  addText = FALSE,
  addText_modal = FALSE
) +
  scale_fill_manual(values = c("darkred", "darkblue")) +
  scale_color_manual(values = c("darkred", "darkblue")) +
  labs(title = "") +
  theme_hc()

ggInterval_radarplot(
  Environment,
  plotPartial = c(4, 6),
  showLegend = FALSE,
  base_circle = FALSE,
  base_lty = 1,
  addText = FALSE,
  addText_modal = FALSE,
  type = "rect"
) +
  scale_fill_manual(values = c("darkred", "darkblue")) +
  scale_color_manual(values = c("darkred", "darkblue")) +
  labs(title = "") +
  theme_hc()

ggInterval_radarplot(
  facedata,
  base_circle = FALSE,
  base_lty = 1,
  type = "quantile",
  quantileNum = 5,
  showLegend = TRUE,
  Drift = 0
) +
  scale_fill_brewer(palette = "Greys") +
  labs(title = "", fill = "Quantiles") +
  theme_hc()
```

# Customization

```{r customization_examples}
myIris <- classic2sym(iris, groupby = "Species")

ggInterval_indexplot(myIris$intervalData, aes(x = Sepal.Length, fill = "steelblue")) +
  labs(title = "Customized Index Plot using K-Means Results", fill = "Species") +
  theme_minimal()
```

# Working with Other SDA Packages

## Histogram-Valued Data via HistDAWass

This example is evaluated only when the optional package `HistDAWass` is
installed.

```{r histdawass_example, eval = requireNamespace("HistDAWass", quietly = TRUE)}
BLOOD <- HistDAWass::BLOOD
blood.min <- HistDAWass::get.MatH.stats(BLOOD, stat = "min")
blood.max <- HistDAWass::get.MatH.stats(BLOOD, stat = "max")
blood <- data.frame(blood.min, blood.max)

myBLOOD <- classic2sym(
  blood,
  groupby = "customize",
  minData = blood[, 2:4],
  maxData = blood[, 6:8]
)

colnames(myBLOOD$intervalData) <- HistDAWass::get.MatH.main.info(BLOOD)$varnames
ggInterval_PCA(myBLOOD$intervalData, plot = FALSE)$ggplotPCA
```

## Midpoint-Range Data via MAINT.Data

This example is evaluated only when the optional package `MAINT.Data` is
installed.

```{r maint_data_example, eval = requireNamespace("MAINT.Data", quietly = TRUE)}
AbaloneIdt <- MAINT.Data::AbaloneIdt

AbaloneRange <- exp(AbaloneIdt@LogR)
AbaloneMid <- AbaloneIdt@MidP
AbaloneBounds <- data.frame(
  AbaloneMid - AbaloneRange / 2,
  AbaloneMid + AbaloneRange / 2
)

myAbalone <- classic2sym(
  AbaloneBounds,
  groupby = "customize",
  minData = AbaloneBounds[, 1:7],
  maxData = AbaloneBounds[, 8:14]
)

colnames(myAbalone$intervalData) <- AbaloneIdt@VarNames
ggInterval_PCA(myAbalone$intervalData, plot = FALSE)$ggplotPCA
```

# PCA and Correlation Heatmaps

```{r pca_example}
ggInterval_PCA(
  facedata,
  concepts_group = Subjects,
  showLabels = TRUE,
  labelSize = 3,
  checkOverlap = TRUE
)$ggplotPCA +
  scale_fill_brewer(palette = "Set1") +
  coord_fixed(ratio = 1)
```

```{r corrplot_example}
ggInterval_corrplot(facedata, method = "BG", triangle = "lower")
```

# Citation

Use `citation("ggInterval")` after installation to retrieve the current
package citation metadata.

# Session Information

```{r session_info}
sessionInfo()
```
